Notes of Basic Topolgy

### 2016-11-27 Basic Topology Notes/sections/ch5-fundamental-groups.tex
Adding all recent diagrams and pictures for Basic topology note.

By a \nomen{loop} we mena a continuous map $\alpha:I\to X$ such that
$\alpha(0)=\alpha(1)$. We can view is also as a continuous map
$\alpha:S^1 \to X$. It is said to be based at the point
$\alpha(0)$. Two loops $\alpha$ and $\beta$ with the same \nomen{base
point} can be multiplied, and their product is defined on page 87.
A visualization is here:
% cd 1 todo

But this product is not sufficient to become a group. At least, the
multiplication is not associative:
% cd 2 todo

But clearly the two result are exactly if we do care how long they
occupy on the interval $I$, if the interval $I$ is considered as a
time parameter. So we define the following homotopy relation between
loops. If we can find a family $\{f_r\}$ of maps, one for each $r\in
[0,1]$, such that $f_0=\alpha$, $f_1=\beta$, then we say that the
loops $\alpha$ and $\beta$ are homotopic. Schematically,
% cd 3 todo

This relation can be generalized to any continuous maps:
\begin{defi}[Homotopic]
\nomenclature{Homotopic}{\nomrefpage.}
    Let $f,g: X \to Y$ be continuous maps. Then $f$ is homotopic to
    $g$ if there exists a map $F:X \times I\to Y$ such that $F(x,O) =
    f(x)$ and $F(x,1) = g(x)$ for alt points $x\in X$.
\end{defi}

The map $F$ is called a \nomen{homotopy} from $f$ to $g$, and we write
\nomen{$f\simeq_F g$}. In addition, if $f$ and $g$ agree on some
$A\subset X$, we may wish to deform $f$ to $g$ without altering the
values of $f$ on $A$. In this case we ask for a homotopy $F$ from $f$
to $g$ with the additional property that 
\begin{equation}
    F(a,t)=f(a) \text{ for all $a\in A$, for all $t\in I$}
\end{equation}
when such a homotopy exists, we say the \nomen{$f$ is homotopic to $g$
relative to $A$} and write \nomen{$f\simeq_F g$ rel $A$}.
% cd 4 todo

@@

% fig 5.1 p 88 todo

\begin{ex}
    The author shows on page 88 of \cite{book} that: when $C$ is a
    convex subset of a euclidean space, let $f,g:X\to C$ be continuous
    maps, then $f\simeq_F g$, where $F$ is $F(x,t)=(1-t)f(x)+tg(x)$.
    Note that if $f$ and $g$ agree on a subset $A$ of $X$, then this
    homotopy is a homotopy relative to $A$. This $F$ is called a
    \nomen{straight-line homotopy}.
\end{ex}
\begin{ex}
    Let $f,g:X\to S^n$ be continuous maps. We can take $S^n$ to be the
    unit sphere in $\mathbb{E}^{n+1}$, and think of $f,g$ as
    continuous maps into $\mathbb{E}^{n+1}$, then we may form a
    straight-line homotopy from $f$ to $g$ by:
    \begin{equation}
        F(x,t) = \frac{(1-t)f(x)+tg(x)}{||(1-t)f(x)+tg(x)||}
    \end{equation}
    But why we are  % yes why! TODO
\end{ex}
\begin{ex}
    This is example is best illustrated by pictures:
    % fig 5.2 todo
    Geometrieally, $\alpha$ winds eaeh of the segments
    $[O,\frac{1}{2}]$, $[\frac{1}{2},\frac{3}{4}]$, $[\frac{3}{4}, 1]$
    once round the eirc1e, the first two being wound in an
    anticlockwise direetion, and the third clockwise. The loop $\beta$
    simply winds the whole interval $[0,1]$ once round the circle
    anticlockwise.

@@

    We need another lemma to prove that it is injective.
    \begin{lemma}[Homotopy-lifting lemma]
        If $F:[0,1]\times[0,1]\to S^1$ is a map such that
        $F(0,t)=F(1,t)=1$ for $0\leq t\leq 1$, then there exists a
        unique $\widetilde F: [0,1]\times[0,1]\to \mathbb{R}$ such that
        \begin{align}
            \pi\circ \widetilde F = F \\
            \widetilde F(0,t) =0,\, 0\leq t\leq 1
        \end{align}
    \end{lemma}
    % TODO Graph here. DC 1
    \begin{proof}
        The proof is on page 98 of \cite{book}. The idea is exactly
        the same as in lemma~\ref{thm:path-lifting-lemma-1}.
    \end{proof}
    Now we proof that the map $\phi$ is injective. Suffice to prove
    that $\Ker(\phi)$ is trivial. Suppose $\phi(n)=\pi\circ \gamma_n$
    is homotopic to the constant loop. Then choose a homotopy $F$
    from $\pi\circ\gamma_n$ to the constant loop. By the
    homotopy-lifting lemma we can find $\widetilde F: [0,1]\times[0,1]\to
    \mathbb{R}$ such that $\pi\circ \widetilde F = F$. Also
    $\widetilde F(0,t)=0$.
    We can find the vertical bottom is $0$ and verticl top is
    $\gamma_n$. Right line is integers and can only be $0$ Hence
    $\gamma_n$ starts at $0$ and ends at $0$.
    One can find that $\gamma_n\cong 0$. This completes the proof of
    injectivity. Hence completes the whole proof.
\end{proof}

We have an application,
\begin{thm}[Brouwer Fixed Point theorem] % TODO brower? broer?
    A contiuous map $f:B^2\to B^2$ ($B^2:2D$-closed dicks) must have a
fixed point. That is, $\exists x\in B^2$ such that $f(x)=x$.
\end{thm}
\begin{proof}
    Assuming that this theorem is false, that is $\forall x\in B^2$,
    $x\neq f(x)$, then we have a straight path from $f(x)$ to $x$. We can
    extends this path to cuts the boundary of $B^2$ at $h(x)$. This is
    for all $x\in B^2$, hence we have a map $h: B^2 \to S^1$. Also,
    $h|_{S^1}$ is obvious an identity map. But $S^1$ can be included
    inside $B^2$, so we have:
    \begin{equation}
        S^1 \to B^2 \to S^1 % TODO add function labels
    \end{equation}
    Hence we have a series of homomorphism of fundamental groups:
    \begin{equation}
        \pi_1 (S^1) \to \pi_1 (B^2) \to \pi_1(S^1)
    \end{equation}
    and the composite is identity map. But observe that $B^2$ is a
    convex set and hence its fundamental group is trivial. But $S^1$
    has non-trivial fundamental group.
\end{proof}
\begin{remark}
    This theorem can be extended to higher dimensional case. But the
    proof cannot be the same because for higher dimension $\pi_1(S^n)$
    is no longer non-trivial.
\end{remark}

>>>>

By a \nomen{loop} we mena a continuous map $\alpha:I\to X$ such that
$\alpha(0)=\alpha(1)$. We can view is also as a continuous map
$\alpha:S^1 \to X$. It is said to be based at the point
$\alpha(0)$. Two loops $\alpha$ and $\beta$ with the same \nomen{base
point} can be multiplied, and their product is defined on page 87.
A visualization of $\alpha\cdot\beta$ is here:
\begin{figure}[H]
    \centering
    \includegraphics[width=0.4\linewidth]{pics/multiply-loops.pdf}
    \caption{Multiply loops}
\end{figure}
But this product is not sufficient to become a group. At least, the
multiplication is not associative:
\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{pics/multiply-loop-not-associative.pdf}
    \caption{Multiplication of loop is not associative}
\end{figure}

But clearly the two result are exactly if we do care how long they
occupy on the interval $I$, if the interval $I$ is considered as a
time parameter. So we define the following homotopy relation between
loops. If we can find a family $\{f_r\}$ of maps, one for each $r\in
[0,1]$, such that $f_0=\alpha$, $f_1=\beta$, then we say that the
loops $\alpha$ and $\beta$ are homotopic. Schematically,
\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{pics/homotopy-for-loops.pdf}
    \caption{Homotopy for loops}
\end{figure}
This relation can be generalized to any continuous maps:
\begin{defi}[Homotopic]
\nomenclature{Homotopic}{\nomrefpage.}
    Let $f,g: X \to Y$ be continuous maps. Then $f$ is homotopic to
    $g$ if there exists a map $F:X \times I\to Y$ such that $F(x,O) =
    f(x)$ and $F(x,1) = g(x)$ for alt points $x\in X$.
\end{defi}

The map $F$ is called a \nomen{homotopy} from $f$ to $g$, and we write
\nomen{$f\simeq_F g$}. In addition, if $f$ and $g$ agree on some
$A\subset X$, we may wish to deform $f$ to $g$ without altering the
values of $f$ on $A$. In this case we ask for a homotopy $F$ from $f$
to $g$ with the additional property that 
\begin{equation}
    F(a,t)=f(a) \text{ for all $a\in A$, for all $t\in I$}
\end{equation}
when such a homotopy exists, we say the \nomen{$f$ is homotopic to $g$
relative to $A$} and write \nomen{$f\simeq_F g$ rel $A$}.
\begin{figure}[H]
    \centering
    \includegraphics[width=0.5\linewidth]{pics/homotopy-schematic.pdf}
    \caption{Homotopy Scheme}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{pics/homotopy-for-loops-2.PNG}
    \caption{Homotopy for loops 2 (pp. 88 of \cite{book})}
\end{figure}
\begin{ex}
    The author shows on page 88 of \cite{book} that: when $C$ is a
    convex subset of a euclidean space, let $f,g:X\to C$ be continuous
    maps, then $f\simeq_F g$, where $F$ is $F(x,t)=(1-t)f(x)+tg(x)$.
    Note that if $f$ and $g$ agree on a subset $A$ of $X$, then this
    homotopy is a homotopy relative to $A$. This $F$ is called a
    \nomen{straight-line homotopy}.
\end{ex}
\begin{ex}
    Let $f,g:X\to S^n$ be continuous maps. We can take $S^n$ to be the
    unit sphere in $\mathbb{E}^{n+1}$, and think of $f,g$ as
    continuous maps into $\mathbb{E}^{n+1}$, then we may form a
    straight-line homotopy from $f$ to $g$ by:
    \begin{equation}
        F(x,t) = \frac{(1-t)f(x)+tg(x)}{||(1-t)f(x)+tg(x)||}
    \end{equation}
    But why we are  % yes why! TODO
\end{ex}
\begin{ex}
    This is example is best illustrated by pictures:
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.8\linewidth]{pics/{loops-to-loops-fig5.2}.PNG}
        \caption{Loops to loops (pp. 89 of \cite{book})}
    \end{figure}
    Geometrieally, $\alpha$ winds eaeh of the segments
    $[O,\frac{1}{2}]$, $[\frac{1}{2},\frac{3}{4}]$, $[\frac{3}{4}, 1]$
    once round the eirc1e, the first two being wound in an
    anticlockwise direetion, and the third clockwise. The loop $\beta$
    simply winds the whole interval $[0,1]$ once round the circle
    anticlockwise.

@@

    We need another lemma to prove that it is injective.
    \begin{lemma}[Homotopy-lifting lemma]
        If $F:[0,1]\times[0,1]\to S^1$ is a map such that
        $F(0,t)=F(1,t)=1$ for $0\leq t\leq 1$, then there exists a
        unique $\widetilde F: [0,1]\times[0,1]\to \mathbb{R}$ such that
        \begin{align}
            \pi\circ \widetilde F &= F \\
            \widetilde F(0,t) &=0,\, 0\leq t\leq 1
        \end{align}
    \end{lemma}
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.5\linewidth]{pics/homotopy-lifting-1.pdf}
        \caption{Homotopy lifting lemma-1}
    \end{figure}
    \begin{proof}
        The proof is on page 98 of \cite{book}. The idea is exactly
        the same as in lemma~\ref{thm:path-lifting-lemma-1}.
    \end{proof}
    Now we proof that the map $\phi$ is injective. Suffice to prove
    that $\Ker(\phi)$ is trivial. Suppose $\phi(n)=\pi\circ \gamma_n$
    is homotopic to the constant loop. Then choose a homotopy $F$
    from $\pi\circ\gamma_n$ to the constant loop. By the
    homotopy-lifting lemma we can find $\widetilde F: [0,1]\times[0,1]\to
    \mathbb{R}$ such that $\pi\circ \widetilde F = F$. Also
    $\widetilde F(0,t)=0$.
    We can find the vertical bottom is $0$ and verticl top is
    $\gamma_n$. Right line is integers and can only be $0$. 
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.6\linewidth]{pics/homotopy-lifting-2.pdf}
        \caption{Schematic Draft}
    \end{figure}
    Hence $\gamma_n$ starts at $0$ and ends at $0$.  One can find that
    $\gamma_n\cong 0$. This completes the proof of injectivity. Hence
    completes the whole proof.
\end{proof}

We have an application,
\begin{thm}[Brouwer Fixed Point theorem]
    A contiuous map $f:B^2\to B^2$ ($B^2:2D$-closed dicks) must have a
fixed point. That is, $\exists x\in B^2$ such that $f(x)=x$.
\end{thm}
\begin{proof}
    Assuming that this theorem is false, that is $\forall x\in B^2$,
    $x\neq f(x)$, then we have a straight path from $f(x)$ to $x$. We can
    extends this path to cuts the boundary of $B^2$ at $h(x)$. This is
    for all $x\in B^2$, hence we have a map $h: B^2 \to S^1$. Also,
    $h|_{S^1}$ is obvious an identity map. But $S^1$ can be included
    inside $B^2$, so we have:
    \begin{equation}
        S^1 \to B^2 \to S^1
    \end{equation}
    Hence we have a series of homomorphism of fundamental groups:
    \begin{equation}
        \pi_1 (S^1) \to \pi_1 (B^2) \to \pi_1(S^1)
    \end{equation}
    and the composite is identity map. But observe that $B^2$ is a
    convex set and hence its fundamental group is trivial. But $S^1$
    has non-trivial fundamental group. It is then impossible to form
    such a chain of homomorphism whose product is identity map.
    Contradiction!
\end{proof}
\begin{remark}
    This theorem can be extended to higher dimensional case. But the
    proof cannot be the same because for higher dimension $\pi_1(S^n)$
    is no longer non-trivial.
\end{remark}
